Predicting Water Treatment Plant Chemical Doses

### 2025-12-11 Presentation DSCI445.Rmd
Final final updates

**One role of an operator in the water treatment process is to look at a wide array of data and make decisions about how to adjust chemical doses.**

@@

**OUR GOAL: Create a model that serves the role of an "operator," taking raw water data and making predictions of chemical doses.**

```{r}
img_array = readPNG("Images/treatmentplot3.png")
grid.raster(img_array)
```

@@

- Raw Water Data
   - pH
   - Temperature (of the water)
   - Turbidity
   - Conductivity
   - Suspended Grain Size
   - Alkalinity
   - Hardness

- Chemical Dosing Data
   - Ferric Chloride Dose: primary additive
   - Cationic Polymer Dose: secondary additive

# Data Cleaning

- Dropped Data  
   - Suspended grain size was not used as it was highly variable
   - Alkalinity and hardness were not used because of missing data chunks and differing time ranges

- Processing
   - Daily averages were taken for raw water data to match the dosing data's timestep
   - Significant outliers (likely caused by equipment malfunction) were removed or interpolated
   - Dosing and raw water data merged

@@

Temperature and pH seem to have the strongest relationship to dose, consistent with feedback from treatment plant operators. Other data were either highly variable or had minimum availability with large gaps. We will focus on the following to predict dose:

- pH
- Temperature
- Conductivity (lots of noise, weaker)
- Turbidity (lots of noise, weaker)

Focus: **FeCl Dose**. Polymer dose is not changed significantly in plants.

@@

Similar to the stock dataset we've seen in labs and homeworks, our time series dataset needs **lagged variables**.

- For each of the main four parameters:
   - Introduce two lagged variables
   - First represents the day before
   - Second represents two days before

We also introduced a **categorical predictor** indicating whether the dose increased, decreased, or stayed the same compared to the day before.

@@

- Training and Testing
   - Training 2018-2019, Testing 2020
   - Approximately 70/30 split
- Coagulant Dose Prediction
   - Linear Regression Models
   - Tree-Based Models
- Coagulant Dose Change Prediction
   - LDA, QDA
   - Multinomial
   - Random Forest

>>>>

**An operator looks at a wide array of data and makes decisions about chemical dosing.**

**OUR GOAL: Create a model that serves as an "operator," taking raw water data and making predictions of chemical doses.**

@@

@@

- **Raw Water Data**
   - pH
   - Temperature (of the water)
   - Turbidity
   - Conductivity
   - Suspended Grain Size (dropped)
   - Alkalinity (dropped)
   - Hardness (dropped)
- **Chemical Dosing Data**
   - Ferric Chloride Dose: primary
   - Cationic Polymer Dose: secondary
- Data Cleaning
   - Removed significant outliers
   - Took daily averages for raw water data
   - Merged common dates between datasets

@@

Based on preliminary analyses, we will focus on the following dose predictors:

- **pH** (strongest relationship)
- **Temperature** (strongest relationship)
- **Conductivity** (some noise, weaker)
- **Turbidity** (lots of noise, weaker)

Outcome: **FeCl Dose**. (Polymer dose is not changed significantly in plants.)

@@

Introduced **lagged variables** like in the Stock dataset from the homework.

- For each of the main four variables:
   - Introduce two lagged variables
   - First represents the day before
   - Second represents two days before

Introduced a **categorical predictor** indicating whether the dose increased, decreased, or stayed the same compared to the day before.

@@

- Training and Testing
   - Training 2018-2019, Testing 2020
   - Approximately 70/30 split
- **Coagulant Dose Prediction**
   - Linear Regression Models
   - Tree-Based Models
- **Coagulant Dose Change Prediction**
   - LDA, QDA Classification
   - Multinomial Classification
   - Random Forest Classification
